## R code for analysis of Thailand Isotope Data.

Cyler Conrad, Department of Anthropology, University of New Mexico, cylerc@unm.edu

Charlotte L. King, Department of Anatomy, University of Otago, charlotte.king@otago.ac.nz

Ben Marwick, Department of Anthropology, University of Washington, bmarwick@uw.edu

This document contain R code to reproduce the plots and statistical analysis presented in: 

> Conrad, C., King, C.L. and Marwick, B. (in preparation) Oxygen (δ18O) Isotope Variability in Human Tooth Enamel from Northeast Thailand. Archaeological Research in Asia. 

All data required to perform the analyses can be found at the XXXX digital electronic repository (XXXX). The development version of this document can be found at https://github.com/cylerc/ARA_TI

Details of the computational environment and software dependencies are listed at the end of this document. 

References: 
XXXX

```{r setup}
# set the base directory for knitr to the directory above this one
# install.packages(c("car","devtools", ggmap", "ggplot2", "siar"))
library(knitr)
opts_knit$set(root.dir = '../', progress = FALSE)
```

```{r load_libraries, message = FALSE, warning = FALSE}
# see the output of sessionInfo() at the bottom for package version numbers
library(car)
library(devtools)
library(ggmap)
library(ggplot2) 
library(siar)
```

```{r load dataset}
# Data is transcribed from published literature (see Conrad et al. in preparation) and is uploaded here from a source .csv file.
thai <- read.csv("Data/Thai_Human_Iso.csv")
thai
```

First, plot all sites in NE Thailand. 

```{r map sites}
# Create matrix to plot sites in NE Thailand. Decimal latitude and longitude from published literature (see Conrad et al. in preparation).
sites <-read.table(text="
  id lat lon
  BC 17.40638889 103.2397222
  BLK 15.24888889 102.3430556
  NUL 15.24888889 102.3430556
  KPD 13.41666667 101.1666667
  BNW 15.26722222 102.2763889", header=TRUE)
sites
map <- get_map(location = 'Thailand', zoom = 7)
ggmap(map) + 
  geom_point(data=sites, aes(x = lon, y = lat),alpha = .8, color="darkred", size = 5) + 
  geom_text(data = sites, aes(x = lon, y = lat, label = id), size = 4, vjust = 0, hjust = -0.5, face="bold", position = position_jitter(w = 0.0, h = 0.5)) +
  xlab("Longitude") + 
  ylab("Latitude") + 
  labs(title="Sites in Northeast Thailand")

#Because BNW, NUL and BLK are so close together, the automated commands in R to label these points do not work well. We have aggregated these sites together as one point for our revised map. 

sitesrevised <- sites <-read.table(text="
  id lat lon
  BC 17.40638889 103.2397222
  KPD 13.41666667 101.1666667
  BNW/NUL/BLK 15.26722222 102.2763889", header=TRUE)
sitesrevised
map <- get_map(location = 'Thailand', zoom = 7)
ggmap(map) + 
  geom_point(data=sitesrevised, aes(x = lon, y = lat), color="darkred", size = 5) +
  geom_text(data = sites, aes(x = lon, y = lat, label = id), size = 4, vjust = 0, hjust = -0.3, face="bold") +
  xlab("Longitude") + 
  ylab("Latitude") + 
  labs(title="Sites in Northeast Thailand")
fig_width <- 300 # play with this number and save as high resolution png
ggsave(filename = "Figures/Map_revised.png",
       dpi = 600, units = "mm",
       height = fig_width/1.6, width =  fig_width)
```

Then, plot all human tooth enamel stable isotope data. 

``` {r plot isotope data}
#1. Create plot of all enamel apatite data. Oxygen to Carbon. 
cols <- c("1" = "red","2" = "black","3" = "green", "4" = "orange", "5" = "blue")
ggplot(data=thai, aes(thai$X18o, thai$X13c, colour = factor(thai$Group))) +
  geom_point(size=5, alpha = 0.75) +
  xlab(expression(paste(delta^{18},'O'))) +
  ylab(expression(paste(delta^{13},'C'))) +
  xlim(23,29) +
  ylim(-16,-6) +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        strip.background = element_rect(colour="white", fill="white"),
        strip.text.x = element_text(colour = "black", size = 15),
        axis.title.x = element_text(face="bold", colour="black", size=20),
        axis.text.x  = element_text(colour="black", vjust=0.5, size=15),
        axis.title.y = element_text(face="bold", colour="black", size=20),
        axis.text.y  = element_text(colour="black", vjust=0.5, size=15),
        legend.text = element_text(size=15), 
        legend.title = element_text(size=15)) +
  scale_colour_manual(name = "Site", values = cols, breaks = c("1", "2", "3", "4", "5"),
                      labels = c("Ban Chiang", "Ban Lum Khao",
                                 "Noen U-Loke", "Ban Non Wat", "Khok Phanom Di"))
fig_width <- 300 
ggsave(filename = "Figures/Oxygen-Carbon.png",
       dpi = 600, units = "mm",
       height = fig_width/1.6, width =  fig_width)

#Carbon to Oxygen
cols <- c("1" = "red","2" = "black","3" = "green", "4" = "orange", "5" = "blue")
ggplot(data=thai, aes(thai$X13c, thai$X18o, colour = factor(thai$Group))) +
  geom_point(size=5, alpha = 0.75) +
  xlab(expression(paste(delta^{13},'C'))) +
  ylab(expression(paste(delta^{18},'O'))) +
  xlim(-16,-6) +
  ylim(23,29) +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        strip.background = element_rect(colour="white", fill="white"),
        strip.text.x = element_text(colour = "black", size = 15),
        axis.title.x = element_text(face="bold", colour="black", size=20),
        axis.text.x  = element_text(colour="black", vjust=0.5, size=15),
        axis.title.y = element_text(face="bold", colour="black", size=20),
        axis.text.y  = element_text(colour="black", vjust=0.5, size=15),
        legend.text = element_text(size=15), 
        legend.title = element_text(size=15)) +
  scale_colour_manual(name = "Site", values = cols, breaks = c("1", "2", "3", "4", "5"),
                      labels = c("Ban Chiang", "Ban Lum Khao",
                                 "Noen U-Loke", "Ban Non Wat", "Khok Phanom Di"))
fig_width <- 300 
ggsave(filename = "Figures/Carbon-Oxygen.png",
       dpi = 600, units = "mm",
       height = fig_width/1.6, width =  fig_width)
```

Use Bayesian standard ellipse areas to explore differences between groups.

```{r SIBER standard ellipse areas - all}
#1. Plotting standard ellipse areas for all sites. Colors match previous plots. 
rm(list = ls())
graphics.off()
#Reload data. 
thai <- read.csv("Data/Thai_Human_Iso.csv")
thai
ngroups <- length(unique(thai$Group))
spx <- split(thai$X18o, thai$Group)
spy <- split(thai$X13c, thai$Group)
SEA <- numeric(ngroups)
SEAc <- numeric(ngroups)
TA <- numeric(ngroups)
colorgroup = c(rep("red",length(which(thai$Group==1)))
               ,rep("black",length(which(thai$Group==2)))
               ,rep("green",length(which(thai$Group==3)))
               ,rep("orange",length(which(thai$Group==4)))
               ,rep("blue",length(which(thai$Group==5))))
pchgroup = c(rep(16,length(which(thai$Group==1)))
             ,rep(16,length(which(thai$Group==2)))
             ,rep(16,length(which(thai$Group==3)))
             ,rep(16,length(which(thai$Group==4)))
             ,rep(16,length(which(thai$Group==5))))
png('Figures/StandardEllipseArea.png')
plot(thai$X18o,thai$X13c,col=colorgroup,type="p",xlim=c(24,29),ylim=c(-16,-6),axes=F,pch=pchgroup, cex=1.5, xlab = expression(~ delta ^"18"~'O'), ylab = expression(~ delta^"13"~'C'))
legend("topright",col=c("red", "black", "green", "orange", "blue"),c("Ban Chiang", "Ban Lum Khao", "Noen U-Loke", "Ban Non Wat", "Khok Phanom Di"),bty="o",box.col="black",pch=16,cex=.7)
box()
axis(1,at=seq(24,29,1),labels=T)
axis(2,at=seq(-16,-6,2),labels=F)
axis(2,at=seq(-16,-6,2),labels=T)
for (j in unique(thai$Group)){
  SE <- standard.ellipse(spx[[j]],spy[[j]],steps=1)
  SEA[j] <- SE$SEA
  SEAc[j] <- SE$SEAc
  colorgroup = c('red','black','green','orange','blue')
  lines(SE$xSEAc,SE$ySEAc,lty=1,lwd=3,col=colorgroup[j])
}  
print(cbind(SEA,SEAc))
dev.off()
#We need to considerably increase the dpi and overall size of this plot.
```

Statistically analyze these differences using frequentist statistics. 

```{r all - checking assumptions and comparing means}
#Create data.frames of site data. 
BC <- thai[c(1:42), -1]
BLK <- thai[c(43:68), -1]
NUL <- thai[c(69:102), -1]
BNW <- thai[c(103:219), -1]
KPD <- thai[c(220:285), -1]

##Shapiro-Wilk to test for normality (Sig. = not normal)
#Carbon and oxygen for sites.
shapiro.test(BC$X13c) #not normal
shapiro.test(BC$X18o) #normal
shapiro.test(BLK$X13c) #normal
shapiro.test(BLK$X18o) #not normal
shapiro.test(NUL$X13c) #normal
shapiro.test(NUL$X18o) #normal
shapiro.test(BNW$X13c) #not normal
shapiro.test(BNW$X18o) #normal
shapiro.test(KPD$X13c) #not normal
shapiro.test(KPD$X18o) #normal 

#Test for equal variance between sites. Using Levene's test for groups not following normality (Sig. = unequal variance).
#### Carbon ####
#BC to BLK carbon
data <- c(BC$X13c, BLK$X13c)
site <- c(rep("BC", length(BC$X13c)), rep("BLK", length(BLK$X13c)))
eqv1 <- data.frame(data, site)
eqv1
leveneTest(data ~ site, data = eqv1) #equal

#BC to NUL carbon
data <- c(BC$X13c, NUL$X13c)
site <- c(rep("BC", length(BC$X13c)), rep("NUL", length(NUL$X13c)))
eqv2 <- data.frame(data, site)
eqv2
leveneTest(data ~ site, data = eqv2) #equal

#BC to BNW carbon
data <- c(BC$X13c, BNW$X13c)
site <- c(rep("BC", length(BC$X13c)), rep("BNW", length(BNW$X13c)))
eqv3 <- data.frame(data, site)
eqv3
leveneTest(data ~ site, data = eqv3) #equal

#BC to KPD carbon
data <- c(BC$X13c, KPD$X13c)
site <- c(rep("BC", length(BC$X13c)), rep("KPD", length(KPD$X13c)))
eqv4 <- data.frame(data, site)
eqv4
leveneTest(data ~ site, data = eqv4) #unequal

#BLK to NUL carbon
data <- c(BLK$X13c, NUL$X13c)
site <- c(rep("BLK", length(BLK$X13c)), rep("NUL", length(NUL$X13c)))
eqv5 <- data.frame(data, site)
eqv5
leveneTest(data ~ site, data = eqv5) #equal

#BLK to BNW carbon
data <- c(BLK$X13c, BNW$X13c)
site <- c(rep("BLK", length(BLK$X13c)), rep("BNW", length(BNW$X13c)))
eqv6 <- data.frame(data, site)
eqv6
leveneTest(data ~ site, data = eqv6) #equal

#BLK to KPD carbon
data <- c(BLK$X13c, KPD$X13c)
site <- c(rep("BLK", length(BLK$X13c)), rep("KPD", length(KPD$X13c)))
eqv7 <- data.frame(data, site)
eqv7
leveneTest(data ~ site, data = eqv7) #unequal

#NUL to BNW carbon
data <- c(NUL$X13c, BNW$X13c)
site <- c(rep("NUL", length(NUL$X13c)), rep("BNW", length(BNW$X13c)))
eqv8 <- data.frame(data, site)
eqv8
leveneTest(data ~ site, data = eqv8) #equal

#NUL to KPD carbon
data <- c(NUL$X13c, KPD$X13c)
site <- c(rep("NUL", length(NUL$X13c)), rep("KPD", length(KPD$X13c)))
eqv9 <- data.frame(data, site)
eqv9
leveneTest(data ~ site, data = eqv9) #unequal

#BNW to KPD carbon
data <- c(BNW$X13c, KPD$X13c)
site <- c(rep("BNW", length(BNW$X13c)), rep("KPD", length(KPD$X13c)))
eqv10 <- data.frame(data, site)
eqv10
leveneTest(data ~ site, data = eqv10) #equal 

#### Oxygen ####
#BC to BLK oxygen
data <- c(BC$X18o, BLK$X18o)
site <- c(rep("BC", length(BC$X18o)), rep("BLK", length(BLK$X18o)))
eqv11 <- data.frame(data, site)
eqv11
leveneTest(data ~ site, data = eqv11) #equal

#BC to NUL oxygen
data <- c(BC$X18o, NUL$X18o)
site <- c(rep("BC", length(BC$X18o)), rep("NUL", length(NUL$X18o)))
eqv12 <- data.frame(data, site)
eqv12
leveneTest(data ~ site, data = eqv12) #equal

#BC to BNW oxygen
data <- c(BC$X18o, BNW$X18o)
site <- c(rep("BC", length(BC$X18o)), rep("BNW", length(BNW$X18o)))
eqv13 <- data.frame(data, site)
eqv13
leveneTest(data ~ site, data = eqv13) #unequal 

#BC to KPD oxygen
data <- c(BC$X18o, KPD$X18o)
site <- c(rep("BC", length(BC$X18o)), rep("KPD", length(KPD$X18o)))
eqv14 <- data.frame(data, site)
eqv14
leveneTest(data ~ site, data = eqv14) #unequal

#BLK to NUL oxygen
data <- c(BLK$X18o, NUL$X18o)
site <- c(rep("BLK", length(BLK$X18o)), rep("NUL", length(NUL$X18o)))
eqv15 <- data.frame(data, site)
eqv15
leveneTest(data ~ site, data = eqv15) #unequal 

#BLK to BNW oxygen
data <- c(BLK$X18o, BNW$X18o)
site <- c(rep("BLK", length(BLK$X18o)), rep("BNW", length(BNW$X18o)))
eqv16 <- data.frame(data, site)
eqv16
leveneTest(data ~ site, data = eqv16) #unequal 

#BLK to NUL oxygen
data <- c(BLK$X18o, KPD$X18o)
site <- c(rep("BLK", length(BLK$X18o)), rep("KPD", length(KPD$X18o)))
eqv17 <- data.frame(data, site)
eqv17
leveneTest(data ~ site, data = eqv17) #unequal

#NUL to BNW oxygen
data <- c(NUL$X18o, BNW$X18o)
site <- c(rep("NUL", length(NUL$X18o)), rep("BNW", length(BNW$X18o)))
eqv18 <- data.frame(data, site)
eqv18
leveneTest(data ~ site, data = eqv18) #equal

#NUL to KPD oxygen
data <- c(NUL$X18o, KPD$X18o)
site <- c(rep("NUL", length(NUL$X18o)), rep("KPD", length(KPD$X18o)))
eqv19 <- data.frame(data, site)
eqv19
leveneTest(data ~ site, data = eqv19) #equal

#BNW to KPD oxygen
data <- c(BNW$X18o, KPD$X18o)
site <- c(rep("BNW", length(BNW$X18o)), rep("KPD", length(KPD$X18o)))
eqv20 <- data.frame(data, site)
eqv20
leveneTest(data ~ site, data = eqv20) #equal

#### Testing for difference in means (oxygen). ####
#If not normal use Wilcox, if normal use Welches.
#Welches
#BC to NUL oxygen
t.summary.uneqvar <- t.test(BC$X18o, NUL$X18o, var.equal = TRUE)
t.summary.uneqvar #not significant
#BC to BNW oxygen
t.summary.uneqvar <- t.test(BC$X18o, BNW$X18o, var.equal = FALSE)
t.summary.uneqvar #not significant
#BC to KPD oxygen
t.summary.uneqvar <- t.test(BC$X18o, KPD$X18o, var.equal = FALSE)
t.summary.uneqvar #significant
#NUL to BNW oxygen
t.summary.uneqvar <- t.test(NUL$X18o, BNW$X18o, var.equal = TRUE)
t.summary.uneqvar #significant
#NUL to KPD oxygen
t.summary.uneqvar <- t.test(NUL$X18o, KPD$X18o, var.equal = TRUE)
t.summary.uneqvar #significant
#BNW to KPD oxygen
t.summary.uneqvar <- t.test(BNW$X18o, KPD$X18o, var.equal = TRUE)
t.summary.uneqvar #significant
#Wilcox
wilcox.test(BC$X18o, BLK$X18o, conf.int=TRUE) #significant
wilcox.test(BLK$X18o, NUL$X18o, conf.int=TRUE) #significant
wilcox.test(BLK$X18o, BNW$X18o, conf.int=TRUE) #significant
wilcox.test(BLK$X18o, KPD$X18o, conf.int=TRUE) #not significant

#### Testing for difference in means (carbon). ####
#If not normal use Wilcox, if normal use Welches.
#Welches
#BLK to NUL carbon
t.summary.uneqvar <- t.test(BLK$X13c, NUL$X13c, var.equal = TRUE)
t.summary.uneqvar #significant
#Wilcox
wilcox.test(BC$X13c, BLK$X13c, conf.int=TRUE) #significant
wilcox.test(BC$X13c, NUL$X13c, conf.int=TRUE) #significant
wilcox.test(BC$X13c, BNW$X13c, conf.int=TRUE) #significant
wilcox.test(BC$X13c, KPD$X13c, conf.int=TRUE) #significant 
wilcox.test(BLK$X13c, BNW$X13c, conf.int=TRUE) #significant
wilcox.test(BLK$X13c, KPD$X13c, conf.int=TRUE) #significant
wilcox.test(NUL$X13c, BNW$X13c, conf.int=TRUE) #significant
wilcox.test(NUL$X13c, KPD$X13c, conf.int=TRUE) #significant
wilcox.test(BNW$X13c, KPD$X13c, conf.int=TRUE) #significant
```

##### Now break down data by time period and compare. ##### 

```{r changes through time}
#Create table to guide chronological analysis: mp = mortuary phase
mp <- sites <-read.table(text="
  summary period code site
  early 2100-900BC ep.i-v/mp.i-ii/N1-2/mp.i-vii BC/BLK/BNW/KPD
  middle 900-300BC mp.vi-viii/mp.iii/BA1-5 BC/BLK/BNW
  late 300BC-200AD lp.ix-x/mp.ii-v/IA1-4 BC/NUL/BNW
  ", header=TRUE)
```

Use Bayesian standard ellipse areas to explore differences between groups.

```{r SIBER standard ellipse areas - by time}
#1. Plotting standard ellipse areas for time period. 
rm(list = ls())
graphics.off()
thai <- read.csv("Data/Thai_Human_Iso.csv")
thai
early <- thai[c(1:2,4:6,8:16,22:38,40:65,105:106,110:114,124,133,141:142,160:162,168,175:177,191:192,206:207,209,220:285),]
early
#re-number group column
early$Group <- 1
early
middle <- thai[c(3,7,19:21,39,73:82,103:104,107:109,115,118:123,125:127,129:132,134:140,143:145,148,151:153,155:157,159,163,165:167,169:174,178,182:190,197:198,200:205,208,210:219),]
middle
#re-number group column
middle$Group <- 2
middle
late <- thai[c(17:18,69:102,116:117,128,146:147,149:150,154,158,164,179:181,193:196,199),]
late
#add new group column
late$Group <- 3
late
#bind these new data.frames together
time <- rbind(early, middle, late)
time 
ngroups <- length(unique(time$Group))
spx <- split(time$X18o, time$Group)
spy <- split(time$X13c, time$Group)
SEA <- numeric(ngroups)
SEAc <- numeric(ngroups)
TA <- numeric(ngroups)
colorgroup = c(rep("brown3",length(which(time$Group==1)))
               ,rep("black",length(which(time$Group==2)))
               ,rep("orange3",length(which(time$Group==3))))
pchgroup = c(rep(16,length(which(time$Group==1)))
             ,rep(16,length(which(time$Group==2)))
             ,rep(16,length(which(time$Group==3))))
png('Figures/StandardEllipseArea_Time.png')
plot(time$X18o,time$X13c,col=colorgroup,type="p",xlim=c(24,29),ylim=c(-16,-6),axes=F,pch=pchgroup, cex=1.5, xlab = expression(~ delta ^"18"~'O'), ylab = expression(~ delta^"13"~'C'))
legend("topright",col=c("brown3", "black", "orange3"),c("Early", "Middle", "Late"),bty="o",box.col="black",pch=16,cex=1.1)
box()
axis(1,at=seq(24,29,1),labels=T)
axis(2,at=seq(-16,-6,2),labels=F)
axis(2,at=seq(-16,-6,2),labels=T)
for (j in unique(time$Group)){
  SE <- standard.ellipse(spx[[j]],spy[[j]],steps=1)
  SEA[j] <- SE$SEA
  SEAc[j] <- SE$SEAc
  colorgroup = c('brown3','black','orange3')
  lines(SE$xSEAc,SE$ySEAc,lty=1,lwd=3,col=colorgroup[j])
}  
print(cbind(SEA,SEAc))
dev.off()
```

Statistically analyze these differences using frequentist statistics. 

```{r time - checking assumptions and comparing means}
##Shapiro-Wilk to test for normality (Sig. = not normal)
#carbon and oxygen for groups by time
shapiro.test(early$X13c) #not normal
shapiro.test(early$X18o) #not normal
shapiro.test(middle$X13c) #normal
shapiro.test(middle$X18o) #not normal
shapiro.test(late$X13c) #not normal
shapiro.test(late$X18o) #normal 

#Test for equal variance between group, Levene's for not normal (Sig. = unequal variance)
#Early to middle carbon
data <- c(early$X13c, middle$X13c)
period <- c(rep("Early", length(early$X13c)), rep("Middle", length(middle$X13c)))
etm1 <- data.frame(data, period)
etm1
leveneTest(data ~ period, data = etm1) #unequal 

#Early to late carbon
data <- c(early$X13c, late$X13c)
period <- c(rep("Early", length(early$X13c)), rep("Late", length(late$X13c)))
etm2 <- data.frame(data, period)
etm2
leveneTest(data ~ period, data = etm2) #unequal

#Middle to late carbon
data <- c(middle$X13c, late$X13c)
period <- c(rep("Middle", length(middle$X13c)), rep("Late", length(late$X13c)))
etm3 <- data.frame(data, period)
etm3
leveneTest(data ~ period, data = etm3) #unequal 

#Early to middle oxygen
data <- c(early$X18o, middle$X18o)
period <- c(rep("Early", length(early$X18o)), rep("Middle", length(middle$X18o)))
etm4 <- data.frame(data, period)
etm4
leveneTest(data ~ period, data = etm4) #unequal 

#Early to late oxygen
data <- c(early$X18o, late$X18o)
period <- c(rep("Early", length(early$X18o)), rep("Late", length(late$X18o)))
etm5 <- data.frame(data, period)
etm5
leveneTest(data ~ period, data = etm5) #equal 

#Middle to late oxygen
data <- c(middle$X18o, late$X18o)
period <- c(rep("Middle", length(middle$X18o)), rep("Late", length(late$X18o)))
etm6 <- data.frame(data, period)
etm6
leveneTest(data ~ period, data = etm6) #equal 

#If not normal and unequal variance use Wilcox
wilcox.test(early$X13c, middle$X13c, conf.int=TRUE) #significant
wilcox.test(early$X13c, late$X13c, conf.int=TRUE) #significant
wilcox.test(middle$X13c, late$X13c, conf.int=TRUE) #significant
wilcox.test(early$X18o, middle$X18o, conf.int=TRUE) #significant
wilcox.test(early$X18o, late$X18o, conf.int=TRUE) #significant
wilcox.test(middle$X18o, late$X18o, conf.int=TRUE) #significant
```

##### Now break down data by sexes and compare. #####

```{r SIBER standard ellipse areas - by sex and time}
#Plot male isotope values.
rm(list = ls())
graphics.off()
thai <- read.csv("Data/Thai_Human_Iso.csv")
thai
male <- subset(thai, Sex=="M", select=Group:Notes)
male
ngroups <- length(unique(male$Group))
spx <- split(male$X18o, male$Group)
spy <- split(male$X13c, male$Group)
SEA <- numeric(ngroups)
SEAc <- numeric(ngroups)
TA <- numeric(ngroups)
colorgroup = c(rep("red",length(which(male$Group==1)))
               ,rep("black",length(which(male$Group==2)))
               ,rep("green",length(which(male$Group==3)))
               ,rep("orange",length(which(male$Group==4)))
               ,rep("blue",length(which(male$Group==5))))
pchgroup = c(rep(16,length(which(male$Group==1)))
             ,rep(16,length(which(male$Group==2)))
             ,rep(16,length(which(male$Group==3)))
             ,rep(16,length(which(male$Group==4)))
             ,rep(16,length(which(male$Group==5))))
png('Figures/StandardEllipseArea_Male.png')
plot(male$X18o,male$X13c,col=colorgroup,type="p",xlim=c(24,29),ylim=c(-16,-6),axes=F,pch=pchgroup, cex=1.5, xlab = expression(~ delta ^"18"~'O'), ylab = expression(~ delta^"13"~'C'), main="Male")
legend("topright",col=c("red", "black", "green", "orange", "blue"),c("Ban Chiang", "Ban Lum Khao", "Noen U-Loke", "Ban Non Wat", "Khok Phanom Di"),bty="o",box.col="black",pch=16,cex=.7)
box()
axis(1,at=seq(24,29,1),labels=T)
axis(2,at=seq(-16,-6,2),labels=F)
axis(2,at=seq(-16,-6,2),labels=T)
for (j in unique(male$Group)){
  SE <- standard.ellipse(spx[[j]],spy[[j]],steps=1)
  SEA[j] <- SE$SEA
  SEAc[j] <- SE$SEAc
  colorgroup = c('red','black','green','orange','blue')
  lines(SE$xSEAc,SE$ySEAc,lty=1,lwd=3,col=colorgroup[j])
}  
print(cbind(SEA,SEAc))
dev.off()
#Plot female isotope values. 
rm(list = ls())
graphics.off()
thai <- read.csv("Data/Thai_Human_Iso.csv")
thai
female <- subset(thai, Sex=="F", select=Group:Notes)
female
ngroups <- length(unique(female$Group))
spx <- split(female$X18o, female$Group)
spy <- split(female$X13c, female$Group)
SEA <- numeric(ngroups)
SEAc <- numeric(ngroups)
TA <- numeric(ngroups)
colorgroup = c(rep("red",length(which(female$Group==1)))
               ,rep("black",length(which(female$Group==2)))
               ,rep("green",length(which(female$Group==3)))
               ,rep("orange",length(which(female$Group==4)))
               ,rep("blue",length(which(female$Group==5))))
pchgroup = c(rep(16,length(which(female$Group==1)))
             ,rep(16,length(which(female$Group==2)))
             ,rep(16,length(which(female$Group==3)))
             ,rep(16,length(which(female$Group==4)))
             ,rep(16,length(which(female$Group==5))))
png('Figures/StandardEllipseArea_Female.png')
plot(female$X18o,female$X13c,col=colorgroup,type="p",xlim=c(24,29),ylim=c(-16,-6),axes=F,pch=pchgroup, cex=1.5, xlab = expression(~ delta ^"18"~'O'), ylab = expression(~ delta^"13"~'C'), main="Female")
legend("topright",col=c("red", "black", "green", "orange", "blue"),c("Ban Chiang", "Ban Lum Khao", "Noen U-Loke", "Ban Non Wat", "Khok Phanom Di"),bty="o",box.col="black",pch=16,cex=.7)
box()
axis(1,at=seq(24,29,1),labels=T)
axis(2,at=seq(-16,-6,2),labels=F)
axis(2,at=seq(-16,-6,2),labels=T)
for (j in unique(female$Group)){
  SE <- standard.ellipse(spx[[j]],spy[[j]],steps=1)
  SEA[j] <- SE$SEA
  SEAc[j] <- SE$SEAc
  colorgroup = c('red','black','green','orange','blue')
  lines(SE$xSEAc,SE$ySEAc,lty=1,lwd=3,col=colorgroup[j])
}  
print(cbind(SEA,SEAc))
dev.off()

#### Plot male and female through time ####
#Plot male isotope values through time.
rm(list = ls())
graphics.off()
thai <- read.csv("Data/Thai_Human_Iso.csv")
thai
early <- thai[c(1:2,4:6,8:16,22:38,40:65,105:106,110:114,124,133,141:142,160:162,168,175:177,191:192,206:207,209,220:285),]
early
#re-number group column
early$Group <- 1
early
middle <- thai[c(3,7,19:21,39,73:82,103:104,107:109,115,118:123,125:127,129:132,134:140,143:145,148,151:153,155:157,159,163,165:167,169:174,178,182:190,197:198,200:205,208,210:219),]
middle
#re-number group column
middle$Group <- 2
middle
late <- thai[c(17:18,69:102,116:117,128,146:147,149:150,154,158,164,179:181,193:196,199),]
late
#add new group column
late$Group <- 3
late
#bind these new data.frames together
time <- rbind(early, middle, late)
time 
maletime <- subset(time, Sex=="M", select=Group:Notes)
maletime
ngroups <- length(unique(maletime$Group))
spx <- split(maletime$X18o, maletime$Group)
spy <- split(maletime$X13c, maletime$Group)
SEA <- numeric(ngroups)
SEAc <- numeric(ngroups)
TA <- numeric(ngroups)
colorgroup = c(rep("brown3",length(which(maletime$Group==1)))
               ,rep("black",length(which(maletime$Group==2)))
               ,rep("orange3",length(which(maletime$Group==3))))
pchgroup = c(rep(16,length(which(maletime$Group==1)))
             ,rep(16,length(which(maletime$Group==2)))
             ,rep(16,length(which(maletime$Group==3))))
png('Figures/StandardEllipseArea_Male-Time.png')
plot(maletime$X18o,maletime$X13c,col=colorgroup,type="p",xlim=c(24,29),ylim=c(-16,-6),axes=F,pch=pchgroup, cex=1.5, xlab = expression(~ delta ^"18"~'O'), ylab = expression(~ delta^"13"~'C'), main="Male through Time")
legend("topright",col=c("brown3", "black", "orange3"),c("Early", "Middle", "Late"),bty="o",box.col="black",pch=16,cex=.7)
box()
axis(1,at=seq(24,29,1),labels=T)
axis(2,at=seq(-16,-6,2),labels=F)
axis(2,at=seq(-16,-6,2),labels=T)
for (j in unique(maletime$Group)){
  SE <- standard.ellipse(spx[[j]],spy[[j]],steps=1)
  SEA[j] <- SE$SEA
  SEAc[j] <- SE$SEAc
  colorgroup = c('brown3','black','orange3')
  lines(SE$xSEAc,SE$ySEAc,lty=1,lwd=3,col=colorgroup[j])
}  
print(cbind(SEA,SEAc))
dev.off()
#Plot female isotope values through time. 
rm(list = ls())
graphics.off()
thai <- read.csv("Data/Thai_Human_Iso.csv")
thai
early <- thai[c(1:2,4:6,8:16,22:38,40:65,105:106,110:114,124,133,141:142,160:162,168,175:177,191:192,206:207,209,220:285),]
early
#re-number group column
early$Group <- 1
early
middle <- thai[c(3,7,19:21,39,73:82,103:104,107:109,115,118:123,125:127,129:132,134:140,143:145,148,151:153,155:157,159,163,165:167,169:174,178,182:190,197:198,200:205,208,210:219),]
middle
#re-number group column
middle$Group <- 2
middle
late <- thai[c(17:18,69:102,116:117,128,146:147,149:150,154,158,164,179:181,193:196,199),]
late
#add new group column
late$Group <- 3
late
#bind these new data.frames together
time <- rbind(early, middle, late)
time 
femaletime <- subset(time, Sex=="F", select=Group:Notes)
femaletime
ngroups <- length(unique(femaletime$Group))
spx <- split(femaletime$X18o, femaletime$Group)
spy <- split(femaletime$X13c, femaletime$Group)
SEA <- numeric(ngroups)
SEAc <- numeric(ngroups)
TA <- numeric(ngroups)
colorgroup = c(rep("brown3",length(which(femaletime$Group==1)))
               ,rep("black",length(which(femaletime$Group==2)))
               ,rep("orange3",length(which(femaletime$Group==3))))
pchgroup = c(rep(16,length(which(femaletime$Group==1)))
             ,rep(16,length(which(femaletime$Group==2)))
             ,rep(16,length(which(femaletime$Group==3))))
png('Figures/StandardEllipseArea_Female-Time.png')
plot(femaletime$X18o,femaletime$X13c,col=colorgroup,type="p",xlim=c(24,29),ylim=c(-16,-6),axes=F,pch=pchgroup, cex=1.5, xlab = expression(~ delta ^"18"~'O'), ylab = expression(~ delta^"13"~'C'), main="Female through Time")
legend("topright",col=c("brown3", "black", "orange3"),c("Early", "Middle", "Late"),bty="o",box.col="black",pch=16,cex=.7)
box()
axis(1,at=seq(24,29,1),labels=T)
axis(2,at=seq(-16,-6,2),labels=F)
axis(2,at=seq(-16,-6,2),labels=T)
for (j in unique(femaletime$Group)){
  SE <- standard.ellipse(spx[[j]],spy[[j]],steps=1)
  SEA[j] <- SE$SEA
  SEAc[j] <- SE$SEAc
  colorgroup = c('brown3','black','orange3')
  lines(SE$xSEAc,SE$ySEAc,lty=1,lwd=3,col=colorgroup[j])
}  
print(cbind(SEA,SEAc))
dev.off()
```

Use Bayesian standard ellipse areas to explore differences between sexes.

```{r SIBER standard ellipse areas - by sex}
```

Statistically analyze these differences using frequentist statistics. 

```{r SIBER standard ellipse areas - by sex}
```

##### Bayesian Approaches not completed #####

```{r bayesian statistical analysis}
#1. Use bayesian approaches to understand differences between sites. 
#This needs work here...I'm not entirely sure how to do this in R yet...
```

```{r session_info}
# Computational environment and software dependencies for this analysis.
sessionInfo()
```

LICENSE
 
The MIT License (MIT)

Copyright (c) 2015 Cyler Conrad, Charlotte L. King & Ben Marwick

Permission is hereby granted, free of charge, to any person obtaining a copy
of this software and associated documentation files (the "Software"), to deal
in the Software without restriction, including without limitation the rights
to use, copy, modify, merge, publish, distribute, sublicense, and/or sell
copies of the Software, and to permit persons to whom the Software is
furnished to do so, subject to the following conditions:

The above copyright notice and this permission notice shall be included in
all copies or substantial portions of the Software.

THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT. IN NO EVENT SHALL THE
AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER
LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM,
OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN
THE SOFTWARE.
