## R code for analysis of Thailand Isotope Data.

Cyler Conrad, Department of Anthropology, University of New Mexico, cylerc@unm.edu

Charlotte L. King, Department of Anatomy, University of Otago, charlotte.king@otago.ac.nz

Ben Marwick, Department of Anthropology, University of Washington, bmarwick@uw.edu

This document contain R code to reproduce the plots and statistical analysis presented in: 

> Conrad, C., King, C.L. and Marwick, B. (in preparation) Oxygen (δ18O) Isotope Variability in Human Tooth Enamel from Northeast Thailand. Archaeological Research in Asia. 

All data required to perform the analyses can be found at the XXXX digital electronic repository (XXXX). The development version of this document can be found at https://github.com/cylerc/ARA_TI

Details of the computational environment and software dependencies are listed at the end of this document. 

References: 
XXXX

```{r setup}
# set the base directory for knitr to the directory above this one
# install.packages(c("ggmap", "ggplot2", "siar", "devtools"))
library(knitr)
opts_knit$set(root.dir = '../', progress = FALSE)
```

```{r load_libraries, message = FALSE, warning = FALSE}
# see the output of sessionInfo() at the bottom for package version numbers
library(ggmap)
library(ggplot2) 
library(siar)
library(devtools)
```

```{r load dataset and map sites}
#1. Data is transcribed from published literature (see Conrad et al. in preparation) and is uploaded here from a source .csv file.
thai <- read.csv("Data/Thai_Human_Iso.csv")
thai
#2. Create matrix to plot sites in NE Thailand. Decimal latitude and longitude from published literature (see Conrad et al. in preparation).
sites <-read.table(text="
  id lat lon
  BC 17.40638889 103.2397222
  BLK 15.24888889 102.3430556
  NUL 15.24888889 102.3430556
  KPD 13.41666667 101.1666667
  BNW 15.26722222 102.2763889", header=TRUE)
sites
map <- get_map(location = 'Thailand', zoom = 7)
ggmap(map) + 
  geom_point(data=sites, aes(x = lon, y = lat),alpha = .8, color="darkred", size = 5) + 
  geom_text(data = sites, aes(x = lon, y = lat, label = id), size = 4, vjust = 0, hjust = -0.5, face="bold", position = position_jitter(w = 0.0, h = 0.5)) +
  xlab("Longitude") + 
  ylab("Latitude") + 
  labs(title="Sites in Northeast Thailand")
#Because BNW, NUL and BLK are so close together, the automated commands in R to label these points do not work well. We may have to play with this map more. 
fig_width <- 300 # play with this number and save as high resolution png
ggsave(filename = "Figures/Map.png",
       dpi = 600, units = "mm",
       height = fig_width/1.6, width =  fig_width)
```

``` {r plot isotope data}
#1. Create plot of all enamel apatite data. Oxygen to Carbon. 
cols <- c("1" = "red","2" = "black","3" = "green", "4" = "orange", "5" = "blue")
ggplot(data=thai, aes(thai$X18o, thai$X13c, colour = factor(thai$Group))) +
  geom_point(size=5, alpha = 0.75) +
  xlab(expression(paste(delta^{18},'O'))) +
  ylab(expression(paste(delta^{13},'C'))) +
  xlim(23,29) +
  ylim(-16,-6) +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        strip.background = element_rect(colour="white", fill="white"),
        strip.text.x = element_text(colour = "black", size = 15),
        axis.title.x = element_text(face="bold", colour="black", size=20),
        axis.text.x  = element_text(colour="black", vjust=0.5, size=15),
        axis.title.y = element_text(face="bold", colour="black", size=20),
        axis.text.y  = element_text(colour="black", vjust=0.5, size=15),
        legend.text = element_text(size=15), 
        legend.title = element_text(size=15)) +
  scale_colour_manual(name = "Site", values = cols, breaks = c("1", "2", "3", "4", "5"),
                      labels = c("Ban Chiang", "Ban Lum Khao",
                                 "Noen U-Loke", "Ban Non Wat", "Khok Phanom Di"))
fig_width <- 300 
ggsave(filename = "Figures/Oxygen-Carbon.png",
       dpi = 600, units = "mm",
       height = fig_width/1.6, width =  fig_width)

#Carbon to Oxygen
cols <- c("1" = "red","2" = "black","3" = "green", "4" = "orange", "5" = "blue")
ggplot(data=thai, aes(thai$X13c, thai$X18o, colour = factor(thai$Group))) +
  geom_point(size=5, alpha = 0.75) +
  xlab(expression(paste(delta^{13},'C'))) +
  ylab(expression(paste(delta^{18},'O'))) +
  xlim(-16,-6) +
  ylim(23,29) +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        strip.background = element_rect(colour="white", fill="white"),
        strip.text.x = element_text(colour = "black", size = 15),
        axis.title.x = element_text(face="bold", colour="black", size=20),
        axis.text.x  = element_text(colour="black", vjust=0.5, size=15),
        axis.title.y = element_text(face="bold", colour="black", size=20),
        axis.text.y  = element_text(colour="black", vjust=0.5, size=15),
        legend.text = element_text(size=15), 
        legend.title = element_text(size=15)) +
  scale_colour_manual(name = "Site", values = cols, breaks = c("1", "2", "3", "4", "5"),
                      labels = c("Ban Chiang", "Ban Lum Khao",
                                 "Noen U-Loke", "Ban Non Wat", "Khok Phanom Di"))
fig_width <- 300 
ggsave(filename = "Figures/Carbon-Oxygen.png",
       dpi = 600, units = "mm",
       height = fig_width/1.6, width =  fig_width)
```

```{r SIBER standard ellipse areas}
#1. Plotting standard ellipse areas for all sites. Colors match previous plots. 
rm(list = ls())
#Clear dataframe. 
graphics.off()
#Reload data. 
thai <- read.csv("Data/Thai_Human_Iso.csv")
thai
ngroups <- length(unique(thai$Group))
spx <- split(thai$X18o, thai$Group)
spy <- split(thai$X13c, thai$Group)
SEA <- numeric(ngroups)
SEAc <- numeric(ngroups)
TA <- numeric(ngroups)
colorgroup = c(rep("red",length(which(thai$Group==1)))
               ,rep("black",length(which(thai$Group==2)))
               ,rep("green",length(which(thai$Group==3)))
               ,rep("orange",length(which(thai$Group==4)))
               ,rep("blue",length(which(thai$Group==5))))
pchgroup = c(rep(16,length(which(thai$Group==1)))
             ,rep(16,length(which(thai$Group==2)))
             ,rep(16,length(which(thai$Group==3)))
             ,rep(16,length(which(thai$Group==4)))
             ,rep(16,length(which(thai$Group==5))))
png('Figures/StandardEllipseArea.png')
plot(thai$X18o,thai$X13c,col=colorgroup,type="p",xlim=c(24,29),ylim=c(-16,-6),axes=F,pch=pchgroup, cex=1.5, xlab = expression(~ delta ^"18"~'O'), ylab = expression(~ delta^"13"~'C'))
box()
axis(1,at=seq(24,29,1),labels=T)
axis(2,at=seq(-16,-6,2),labels=F)
axis(2,at=seq(-16,-6,2),labels=T)
for (j in unique(thai$Group)){
  SE <- standard.ellipse(spx[[j]],spy[[j]],steps=1)
  SEA[j] <- SE$SEA
  SEAc[j] <- SE$SEAc
  colorgroup = c('red','black','green','orange','blue')
  lines(SE$xSEAc,SE$ySEAc,lty=1,lwd=3,col=colorgroup[j])
}  
print(cbind(SEA,SEAc))
dev.off()
#It would be good to add a legend to this plot. We also need to considerably increase the dpi and overall size.  
```

```{r frequentist statistical analysis}
#1. Use frequentist approaches to understand differences between sites. 
#Create new matrix of values for individual sites.
BC <- thai[c(1:42), -1]
BLK <- thai[c(43:68), -1]
NUL <- thai[c(69:102), -1]
BNW <- thai[c(103:219), -1]
KPD <- thai[c(220:285), -1]
KPD

#Start with Welch Two Sample t-test.
#BC to BLK
#We expect to see sig. differences between these two sites in d18O space. 
component <-c(BC$X18o, BLK$X18o)
component
Group <- c(rep("One", length(BC$X18o)), rep("Two", length(BLK$X18o)))
group1 <- data.frame(component, Group)
group1
t.summary.uneqvar <- t.test(component ~ Group, data = group1, var.equal = FALSE)
t.summary.uneqvar
#Significantly different at p<0.00001

#BLK and KPD
#We do not expect to see sig. differences between these two sites in d18O space. 
component <-c(BLK$X18o, KPD$X18o)
component
Group <- c(rep("One", length(BLK$X18o)), rep("Two", length(KPD$X18o)))
group2 <- data.frame(component, Group)
group2
t.summary.uneqvar <- t.test(component ~ Group, data = group2, var.equal = FALSE)
t.summary.uneqvar
#Not significantly different at p=0.6036

#...but, what about carbon? These sites should be sig. different in d13C space.
component <-c(BLK$X13c, KPD$X13c)
component
Group <- c(rep("One", length(BLK$X13c)), rep("Two", length(KPD$X13c)))
group3 <- data.frame(component, Group)
group3
t.summary.uneqvar <- t.test(component ~ Group, data = group3, var.equal = FALSE)
t.summary.uneqvar
#Significantly different at p<0.00001 when comparing carbon isotope values. 
```

```{r bayesian statistical analysis}
#1. Use bayesian approaches to understand differences between sites. 
#This needs work here...I'm not entirely sure how to do this in R yet...
```

```{r session_info}
# Computational environment and software dependencies for this analysis.
sessionInfo()
```

LICENSE
 
The MIT License (MIT)

Copyright (c) 2015 Cyler Conrad, Charlotte L. King & Ben Marwick

Permission is hereby granted, free of charge, to any person obtaining a copy
of this software and associated documentation files (the "Software"), to deal
in the Software without restriction, including without limitation the rights
to use, copy, modify, merge, publish, distribute, sublicense, and/or sell
copies of the Software, and to permit persons to whom the Software is
furnished to do so, subject to the following conditions:

The above copyright notice and this permission notice shall be included in
all copies or substantial portions of the Software.

THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT. IN NO EVENT SHALL THE
AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER
LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM,
OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN
THE SOFTWARE.
